---
title: "2BP microbial functional pathways"
output: html_notebook
---

Here we compare the microbial pathways of 2BP vs No-BP patients.

Reference: https://github.com/biobakery/biobakery/wiki/MaAsLin2

# Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

library(Maaslin2)
```

# Load data

Load HUMAnN result.

```{r}
# Pathways
df <- data.table::fread(
  "Inputs_Final/humann_pathabundance_relab.tsv",
  sep="\t",
  header=TRUE,
) %>%
  as.data.frame()
rownames(df) <- df[,"# Pathway"]
```

Load metadata.

```{r}
metadata <- read.csv(
  "Inputs_Final/TA_samples_timepoint.csv"
)

# Relevel 2BP status
metadata$TwoBP_Status <- factor(
  metadata$TwoBP_Status,
  levels=c("No-BP","2BP"))

# Set rownames
rownames(metadata) <- metadata$sample_name
```

Process the pathway file so that it matches with the metadata file, and remove
stratified (ie, species-level) pathways.

```{r}
# Remove the suffix
colnames(df) <- sapply(
  colnames(df), FUN = function(x)
    strsplit(x,"_",fixed=TRUE)[[1]][1]
)

# Retain certain samples
df2 <- df[,metadata$sample_name]

# Remove stratified pathways
df2 <- df2[!grepl("\\|", rownames(df2)),]
```

# Analysis

```{r}
fit_func <- Maaslin2(
  input_data      = df2,
  input_metadata  = metadata, 
  output          = "2BP_maaslin2", 
  fixed_effects   = c("TwoBP_Status"),
  reference       = c("TwoBP_Status,No-BP"),
  min_abundance   = 0,
  min_prevalence  = 0.2,
  analysis_method = "LM",
  normalization   = "NONE",
  transform       = "LOG")
```

Process the MaAsLin2 output for plotting

```{r}
# All pathways
pathway.res <- fit_func$results

# Export
write.csv(
  pathway.res,
  "MaAsLin2_2BP-vs-noBP.csv",
  row.names=FALSE
)

# Pathways with FDR < 0.25
pathway.plot <- pathway.res %>%
  subset(qval<0.25) %>%
  subset(!(feature %in% c("UNMAPPED","UNINTEGRATED")))
```

Clean up the pathways' names

```{r}
# Replace the first ".." with ":"
# Remove the pathway ID, by splitting at the first 
pathway.plot$pathway.name <- sub(
  "..", ": ",
  pathway.plot$feature,
  fixed=TRUE)

# Replace dot around PWY with hyphen
pathway.plot$pathway.name <- gsub(
  ".PWY", "-PWY",
  pathway.plot$pathway.name,
  fixed=TRUE)
pathway.plot$pathway.name <- gsub(
  "PWY.", "PWY-",
  pathway.plot$pathway.name,
  fixed=TRUE)

# Replace remaining dots with space
pathway.plot$pathway.name <- gsub(
  "\\.+", " ",
  pathway.plot$pathway.name)
```

Bar plot. The `coef` column is approximately log2(fold change): https://forum.biobakery.org/t/trying-to-understand-coef-column-and-how-to-convert-it-to-fold-change/3136/9

```{r}
p <- ggplot(data=pathway.plot,
       aes(x=coef,
           y=reorder(pathway.name,coef))) +
  geom_col(fill="#D41159", width=0.6) +
  geom_vline(xintercept = 0, linewidth = 0.4, color = "black") +
  geom_text(
    x=2,
    aes(label=sprintf("Padj = %.2g", qval)),
    hjust=-0.09, size=2.8) +
  scale_x_continuous(limits=c(0,2), expand=c(0,0),
                     breaks=c(0,1,2)) +
  coord_cartesian(clip="off") +
  labs(x="log2(fold change)", y="",
       title="Microbial functional pathways\nassociated with 2BP") +
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=10, face="plain"),
    axis.text.x = element_text(size=10, color="black"),
    axis.text.y = element_text(size=9, color="black"),
    text = element_text(size=10, family="Arial"),
    plot.margin = unit(c(0.5,2,0.5,0), "cm"),
    panel.grid.minor=element_blank(),
    legend.position = "none"
  )
ggsave(
  "microbial_pathway_fdr0.25.svg",
  plot=p,
  width=7.5, height=5.5, units="in"
)
```

